Modern Statistical Methods --- Example Sheet 3

### 2020-12-08 inverse_problems/C7.tex
updaaaaaaaaaaates

\section{Function space priors and Monte Carlo}
Recall that a random variable in $\RR^d$ is called Gaussian if and only if every linear combination of its components is a Gaussian random variable on $\RR$. This can be generalised to any separable Banach space $(\XX, \BB\XX)$: 
\begin{definition}
	Let $\mu$ be a probability measure on $\Prob(\XX, \BB\XX)$ and let $U \sim \mu$. We call $\mu$ \emph{Gaussian} if, for all $\ell \in \XX^*$, the random variable $\ang{\ell, U}$ has a normal distribution.

@@

>>>>

\section{Function space priors and Monte Carlo}
This section covers two different topics: 
\begin{enumerate}
	\item The general definition of a Gaussian distribution on an infinite-dimensional space (useful for constructing priors);
	\item Description of Monte Carlo methods for approximating the posterior. 
\end{enumerate}
\subsection{Gaussian measures}
Recall that a random variable in $\RR^d$ is called Gaussian if and only if every linear combination of its components is a Gaussian random variable on $\RR$. This can be generalised to any separable Banach space $(\XX, \BB\XX)$: 
\begin{definition}
	Let $\mu$ be a probability measure on $\Prob(\XX, \BB\XX)$ and let $U \sim \mu$. We call $\mu$ \emph{Gaussian} if, for all $\ell \in \XX^*$, the random variable $\ang{\ell, U}$ has a normal distribution.

@@

\subsection{Monte Carlo methods}
Generally, Monte Carlo techniques aim at approximating the integral $\bar g \ceq \int_\XX g \dd{\mu}$ by generating independent samples $U_1, U_2, \dotsc \overset{\Rm{iid}}\sim \mu$ and computing the estimator
\[
\bar g \approx \hat g_M \ceq \frac1M \sum_{m=1}^M  g(U_m). 
\]
By the strong law of large numbers, we know that $\hat g_m \to \bar g$ for $M \to\infty$, $\PP$-almost surely. Furthermore, if $\Var_\mu(g)< \infty$, then we have the following expression for the standard deviation:
\[
\sqrt{\EE[(\hat g_m - \bar g)^2]} = \sqrt{\Var_\mu(g)} \cdot M^{-1/2},
\]
which gives a convergence rate of $M^{-1/2}$.

This rate is not the best possible rate: if we assume that $g$ is smooth in some sense, much better rates can be obtained. However, a large advantage is that the rate $M^{-1/2}$ is independent of the dimension of the domain.

The standard Monte Carlo method does not work for Bayesian inverse problems because in general it is not possible to sample independently from the posterior measure. We must therefore sample dependently (there are also other options such as sampling independently from a different measure and correct using weights).

\begin{definition}
	Let $(U_n)_{n=1}^\infty$ be a sequence of $\XX$-valued random variables. Then $(U_n)$ is called a \emph{Markov chain} if, for any $n \in \NN$ and $u_1, \dotsc, u_n \in \XX$, we have
	\[
	\PP(U_{n+1} \in \cdot \mid U_1 = u_1, \dotsc, U_n = u_n) = \PP(U_{n+1} \in \cdot \mid U_n = u_n)
	\]
	We call the chain \emph{time-homogeneous} if, for all $u \in \XX, n \in \NN$ we have
	\[
	 \PP(U_{k+2} \in \cdot \mid U_{k+1} = u) = \PP(U_2 \in \cdot \mid U_1 = u). 
	\]

	A time-homogeneous Markov chain can be fully represented by a Markov kernel $K \colon \BB\XX \times \XX \to [0, 1]$ where
	\[
	K(B \mid u) \ceq \PP(U_2 \in B \mid U_1 = u)
	\]

	Let $\mu \in \Prob(\XX, \BB\XX)$ be any probability meaure, then we denote the \emph{composition} of $\mu$ and $K$ by
	\[
	\mu K(B) \ceq \int_\XX K(B \mid u) \dd{\mu(u)} \qquad(B \in \BB\XX),
	\]
	and we call $\mu$ \emph{stationary} w.r.t.\ $K$ if $\mu K = \mu$. Note that $\mu$ is stationary w.r.t.\ $K$ if and only if $U_1 \sim \mu \implies U_2 \sim \mu$.

	Finally, we say that the Markov kernel $K$ satisfies \emph{detailed balance} w.r.t.\ $\mu' \in \Prob(\XX, \BB\XX)$ if
	\[
	\int_B K(A \mid u) \dd{\mu'(u)} = \int_A K(B \mid u) \dd{\mu'(u)} \qquad(A, B \in \BB\XX). 
	\]
\end{definition}

\begin{lemma}
	Let $K \colon \BB \XX \times \XX \to [0, 1]$ be a Markov kernel that satisfies detailed balance w.r.t.\ $\mu \in \Prob(\XX, \BB\XX)$. Then $K$ is stationary w.r.t.\ $\mu$. 
\end{lemma}

\begin{definition}
	Let $\mu \in \Prob(\XX, \BB\XX, \nu)$ and let $g \colon (\XX, \BB\XX) \to (\RR, \BB\RR)$ be positive with $g = c \dv{\mu}{\nu}$ for some $c > 0$. Moreover, let $Q \colon \XX \times \BB\XX \to [0, 1]$ be a Markov kernel, given by a positive function $q \colon (\XX, \BB\XX)^2 \to (\RR, \BB\RR)$ with
	\[
	Q(A \mid u) = \int_A q(u' \mid u) \dd{\nu(u')} \qquad(A \in \BB\XX, u \in \XX). 
	\]
	Now the \emph{Metropolis-Hastings} Markov kernel is given by
	\[
	K_\Rm{MH}(A \mid u) \ceq \delta(A - u) \int_\XX (1-\alpha(u, u'')) Q(\dd{u''} \mid u) + \int_A \alpha(u, u') Q(\dd{u'} \mid u) \qquad(u \in \XX, A \in \BB\XX),
	\]
	where $\alpha(u, u') \ceq 1 \lor \frac{g(u') q(u \mid u')}{g(u) q(u' \mid u)}$. 
\end{definition}

The \emph{Metropolis-Hastings MCMC method} (MCMC = Markov Chain Monte Carlo) is given as follows:
\begin{enumerate}
	\item Start with some initial $U_1 \in \XX$ and set $m = 1$;
	\item Sample $U^* \sim Q(\cdot \mid U_m)$;
	\item With probability $\alpha(U_m, U^*)$ set $U_{m+1} \leftarrow U^*$, otherwise set $U_{m+1} \leftarrow U_m$. 
	\item Increment $m \leftarrow m + 1$ and go to step 2. 
\end{enumerate}

### 2020-12-08 modern_statistical_methods/C4.tex
updaaaaaaaaaaates

Therefore, we must only understand how we can make $\lambda_j/\hat\tau_j^2$ small. First we will briefly consider its practical use.

@@

>>>>

Therefore, we must only understand how we can make $\lambda_j/\hat\tau_j^2$ small. First we will briefly consider the practical use of the debiased lasso.

@@

\subsubsection{Making the error small}
We will need the following fact about normally distributed random variables:
\begin{lemma} \label{lem:multivariate_normal_conditional}
	Let $\Sigma \in \RR^{p \times p}$ be a covariance matrix with $\Omega \ceq \Sigma^{-1}$.

	Let $x_{-j} \sim N(0, \Sigma_{-j, -j})$ and let $\eps \sim N(0, \Omega_{jj}^{-1})$ where $x_{-j}$ and $\eps$ are independent. Then, defining
	\[
	x_j \ceq x_{-j}\T \gamma^{(j)} + \eps, \qquad\text{where } \gamma^{(j)} \ceq - \Omega_{jj}^{-1} \Omega_{-j, j},
	\]
	we have $x \sim N(0, \Sigma)$. 
\end{lemma}

\begin{proof}
	It is clear that $x$ has a multivariate normal distribution, since it is a linear combination of the independent normal variables $x_{-j}$ and $\eps$. It is therefore enough to compute its mean and covariance. We clearly have
	\[
	\EE[x] = \EE[x_{-j}\T \gamma^{(j)}] + \EE[\eps] = 0.
	\]

	For the variance, let $j = 1$ without loss of generality. 
%	Then we can write by blockwise matrix inversion
%	\[
%	\Sigma = \mqty[ \Sigma_{jj} & \Sigma_{-j, j}\T \\ \Sigma_{-j, j} & \Sigma_{-j, -j}] \implies \Omega = \mqty[ \Omega_{jj}^{-1} + \Omega_{jj}^{-1} \Sigma_{-j, j}\T ]
%	\]
%	
	Then we have
	\[
	\Var(x) = \EE\mqty[x_j^2 & x_j x_{-j}\T  \\
	x_{-j} x_j & x_{-j} x_{-j}\T] \eqqcolon \mqty[ A & B \\ B\T & \Sigma_{-j, -j}]. 
	\]
	We can compute 
	\begin{align*}
		A &= \EE[x_j^2] = \gamma^{(j)\top} \Sigma_{-j, -j} \gamma^{(j)} + \Omega_{jj}^{-1} = \Omega_{jj}^{-1} \qty(1 + \Omega_{jj}^{-1} \Omega_{-j, j}\T \Sigma_{-j, -j} \Omega_{-j, j}) \overset\star= \Sigma_{jj},
	\end{align*}
where $\star$ follows from blockwise matrix inversion.

Analogously, we can compute 
\begin{align*}
	B &= \EE[x_j x_{-j}\T] = \EE[\gamma^{(j)\top} x_{-j} x_{-j}\T] + \EE[\eps] \EE[x_{-j}\T] \\
	&= \gamma^{(j)\top} \Sigma_{-j, -j} = -\Omega_{jj}^{-1} \Omega_{-j, j} \Sigma_{-j, -j} = \Sigma_{j, -j},
\end{align*}
where the last step again follows from blockwise matrix inversion.
This proves that $x \sim N(0, \Sigma)$. 
\end{proof}

By this lemma, we know that the $j$-th variable can be interpreted as the response of a linear model in the other variables with coefficients $\gamma^{(j)}$. Therefore, $\hat\gamma^{(j)}$ defined in \cref{eq:lasso_regress_column} is the Lasso estimator of $\gamma^{(j)}$ with tuning parameter $\lambda_j$. We can use Lasso properties to understand properties of $\hat\gamma^{(j)}$, but for this we need $\gamma^{(j)}$ to be sparse. We define
\[
s \ceq \norm{\beta^0}_0, \quad s_j \ceq \norm{\gamma^{(j)}}_0, \quad s_\Rm{max} \ceq \max(s, s_1, \dotsc, s_p),
\]
where $\norm{v}_0$ is the amount of nonzero entries of $v$.

We assume we have a sequence of models with $n \to \infty$. Recall our setting:
\begin{enumerate}
	\item We have $Y = X\beta^0 + \eps$, where $\eps \sim N(0, \sigma^2I)$. We will assume $\sigma$ is constant. 
	\item We assume that the rows of $X$ are i.i.d.\ $N(0, \Sigma)$ variables, where $\Sigma$ is positive definite with inverse $\Omega$. We let $c$ denote the smallest eigenvalue of $\Sigma$. 
\end{enumerate}
\begin{theorem}
	Suppose $c \geq c_\Rm{min} > 0$ and $\max_j \Sigma_{jj} \leq 1$ for all models in the sequence, and suppose $s_\Rm{max} \sqrt{\log(p) / n} \to 0$. Then there exist constants $A_1, A_2$ such that setting $\lambda = \lambda_j = A_1 \sqrt{\log(p)/n}$ we have
\begin{align*}
	\sqrt n(\hat b - \beta^0) = W + \Delta \qquad\text{where } W \mid X  \sim N(0, \sigma^2 \hat\Theta\Sigma\hat\Theta\T), 
\end{align*}
and as $n, p \to\infty$, 
\[
\PP(\norm{\Delta}_\infty > A_2 s \log(p)/\sqrt n) \to 0. 
\]
\end{theorem}

\begin{proof}
	We define an event $\Lambda_n$ by the following properties:
	\begin{enumerate}
		\item $\phi^2_{\hat\Sigma, s} \geq c_\Rm{min} / 2$ and $\phi^2_{\hat \Sigma_{-j, -j}, s_j} \geq c_\Rm{min} / 2$ for all $j$ (recall the definitions from \cref{thm:lasso_compatibility_condition_asymptotics});

		\item $2\norm{X\T\eps}_\infty / n \leq \lambda$ and $2\norm{X_{-j}\T\eps}_\infty/n \leq \lambda$ for all $j$;
		\item $\norm{\eps^{(j)}}_2^2/n \geq \Omega_{jj}^{-1} (1 - 4\sqrt{\log(p)/n})$ for all $j$.
	\end{enumerate}
On the third example sheet it is shown that $\PP(\Lambda_n) \to 1$ for sufficiently large $A_1$. We work on $\Lambda_n$ and let $c_1, c_2, \dotsc$ be constants.

By the fast rate (\cref{thm:lasso_fast_rate}), we have
\[
\norm{\hat\beta - \beta^0}_1 \leq c_1 s \sqrt{\log(p)/n}. 
\]

Now, fix any $j$, and note that the maximum eigenvalue of $\Omega$ is at most $c_\Rm{min}^{-1}$, so we have
\[
c_\Rm{min}^{-1} \geq \max_{v \in S^{p-1}} v\T \Omega v \geq e_j\T \Omega e_j = \Omega_{jj} \implies  \Omega_{jj}\leq c_\Rm{min}^{-1}. 
\]
Furthermore, we have by \cref{lem:multivariate_normal_conditional} that 
\[
\Omega_{jj}^{-1} = \Var(X_{ij} \mid X_{i, -j}) \overset?\leq \Var(X_{ij}) \leq \max_j \Sigma_{jj} \leq 1.
\]

Applying the fast rate (\cref{thm:lasso_fast_rate}) to the lasso estimator $\hat\gamma^{(j)}$, we have
\begin{align*}
	\norm{\gamma^{(j)} - \hat\gamma^{(j)}}_1 \leq c_2 s_j \Omega_{jj}^{-1} \sqrt{\log(p)/n} \cdot \frac{1}{\phi^2_{\hat\Sigma_{-j, -j}}},
\end{align*}
but since we are working on $\Lambda_n$ we can upper bound the compatibility constant and bound $\Omega_{jj}^{-1} \leq 1$ to obtain
\[
\norm{\gamma^{(j)} -\hat\gamma^{(j)}}_1 \leq c_3 s_j \sqrt{\log(p)/n}.
\]

Now we can lower bound the factor $\hat\tau_j^2$: by definition we have
\begin{align*}
	\hat \tau_j^2 &\geq \frac1n \norm{X_j - X_{-j} \hat\gamma^{(j)}}_2^2 \geq \frac1n \norm{e^{(j)}}_2^2 - \frac2n \norm{X_{-j}\T \eps^{(j)}}_\infty \norm{\gamma^{(j)} - \hat\gamma^{(j)}}_1 \\
	&\overset\star\geq \Omega_{jj}^{-1} (1 - 4 \sqrt{\log(p)/n}) - c_5 \frac{s_\Rm{max} \log(p)}{n} \\
	&\geq c_\Rm{min}(1 - 4 \sqrt{\log(p)/n}) - c_5 \frac{s_\Rm{max} \log p}{n}, 
\end{align*}
and the right hand side goes to 0 for $n \to\infty$ by assumption (here $\star$ follows from the conditions on $\Lambda_n$), so for $n$ large enough we have $\hat\tau_j^2 \geq \frac{c_\Rm{min}}{2}$.

Putting everything together, on $\Lambda_n$ we have
\begin{align*}
	\norm{\Delta}_\infty &\leq \lambda \sqrt n \norm{\hat\beta- \beta^0}_1 \max_j \hat\tau_j^{-2} \\
	&\leq \frac{2}{c_\Rm{min}} \cdot A_1 \sqrt{\log p} \cdot c_1 s \sqrt{\log(p)/n} \\
	&\leq A_2 s \log(p)/\sqrt n, 
\end{align*}
and therefore  $\PP(\norm{\Delta}_\infty > A_2 s \log(p)/\sqrt n) \leq \PP(\Lambda_n\C) \to 0$, which proves the claim.
\end{proof}
